Update of the ozone dry deposition in the OsloCTM3 v1.0

### 2018-10-12 DryDep.tex
Continued writing on Section 2. Started with stomatal and non-stomatal conductance. Included comments by Frode.

\begin{abstract}
  Since the industrial revolution, tropospheric background ozone concentrations have been increasing in the northern hemisphere. In recent years, the number of episodes of peak concentrations has been decreasing in North America and Europe due to the implementation of air quality regulations. At the same time, fast developing countries, like e.g., China or India, saw a significant increase in ozone related air pollution. High concentrations of ozone in ambient air are hazardous not only to humans but to the ecosystem in general. The impact of ozone damage on vegetation and agricultural plants in combination with advancing climate change may affect food security in the future. While the future scenarios in themselves are uncertain, there are limiting factors constraining the accuracy of surface ozone modeling also at present: The distribution and amount of precursor/decomposing species, the stratosphere-troposphere exchange as well as scavenging processes. Removal of any gas through gravitational settling or by uptake by plants and soil is referred to as dry deposition. The process of dry deposition is important for predicting surface ozone concentrations and understanding the observed amount and increase of tropospheric background ozone. The conceptual dry deposition velocities are calculated following a resistance-analogous approach wherein aerodynamic, quasi laminar, and canopy resistances are key components, but these are hard to measure explicitly. In this paper, we present the update from an empirical dry deposition parameterization to a more process based implementation in the OsloCTM3. Our focus lies mainly on a stomatal conductance-based description of the canopy resistance. We evaluate the resulting modeled ozone dry deposition with respect to observations and multi-model studies. We will also estimate the impact of a set of parameters on the modeled ozone concentrations, both at the surface and in the free troposphere. We show that ozone dry deposition decreases with respect to the previous version, leading to an increase in surface ozone of up to ??\,\unit{\%}. {\bf TODO: Finish the results section as soon as there are results.}
\end{abstract}

@@

In the troposphere and in particular in ambient air, ozone is considered as a highly toxic pollutant. Continuously high concentrations of ambient air ozone are hazardous to the whole ecosystem. It is estimated that ozone is cause to an increase in pre-mature deaths \citep{WHO2008}, an average global loss of yield in the four major crops (wheat, rice, maize, and soybean) of about 3--15\,\unit{\%} \citep{PJ:Ainsworth2017} as well as 7\,\unit{\%} loss in primary production in forestry \citep{GCB:Wittig2009,EP:Matyssek2012}. The impact of ozone damage on vegetation and agricultural plants may affect food security in the future especially in Asia \citep{GCB:Tang2013,NCC:Tai2014,AE:Chuwah2015} and might be an important additional feedback to climate change \citep{Nat:Sitch2007}.\\

@@

Removal of any gas through gravitational settling or by uptake by plants and soil is referred to as dry deposition. The process of dry deposition is important for predicting surface ozone concentrations and understanding the observed amount and increase of tropospheric background ozone. It is estimated that about $1000 \pm 200$\,\unit{Tg a^{-1}} of ozone are removed from the atmosphere by dry deposition processes \citep{ACP:Monks2015}. Conceptually, dry deposition is a product between surface ozone concentration \chem{[O_3](h_0)} and a dry deposition velocity $v^\chem{O_3}_d$. Dry deposition velocities $v^i_d$ (also referred to as conductance $G^i$) for any gaseous specie $i$ are typically calculated following a resistance-analogous approach

@@

wherein aerodynamic $R_a$, quasi-laminar layer $R^i_b$, and canopy resistances $R^i_c$ are key components \citep{AE:Wesely1989}. For all gases $R_a$ is the same, while $R^i_b$ and $R^i_c$ vary from gas to gas and also depend on landuse types (e.g., ice/snow, water, urban, desert, agricultural land, deciduous forest, coniferous forest etc.). Originally, \citet{AE:Wesely1989} used fixed seasonal average dry deposition resistances for each landuse type. For all three types of resistances in this Wesely-type parameterization, more process-oriented formulations have been developed and validated over the years. \citet{ACP:Luhar2017} have validated ozone dry deposition to the ocean with respect to three different formulations of surface resistances. An update on the ozone surface resistance over snow and ice covered surfaces has been provided from combined model and observation studies \citep[e.g., $v^\chem{O_3}_\text{ice/snow} = 1/10000\,\unit{cm s^{-1}}$,][]{ACP:Helmig2007}. Canopy conductance is formulated at the single-leave level (stomatal conductance) for various plant function types (PFT) as well as single species based on empirical studies \citep{PTRS:Jarvis1976, BallBerry1987, ACP:Simpson2012, ICP:MappingManual2017}. But progress has also been made on process oriented modeling of stomatal conductance \citep{PP:Buckley2017}. The variety of differing formulations and choices of parameters leads to a wide spread in model intercomparisons \citep{ACP:Hardacre2015} and about 20\,\unit{\%} uncertainty on the resulting total dry deposition \citep{ACP:Monks2015}.\\

@@

\section{Model description}
\label{sec:model_des}
The OsloCTM3 is a three dimensional global chemistry transport model (CTM). The key components of the OsloCTM3 have been described and evaluated by \citet{GMD:Sovde2012}. A detailed account of the capabilities of the OsloCTM3 in simulating anthropogenic aerosol forcing in the past and recent past using the Community Emission Data System (CEDS) historical emission inventory \citep{GMD:Hoesly2018} is given by \citet{GMD:Lund2018}. \emph{The OsloCTM3 can also be coupled to MEGAN \citep{ACP:Guenther2006}. A publication focussing on this is planed.}
Technical: Partly parallelized (low number of effective cores 16, huge memory consumption if run on T159 resolution, code has been ported to fortran90 keeping the dynamical core UCI-CTM mainly intact.)
historical anthropogenic and biomass burning emissions \citep{ACP:Lamarque2010}, meteorological driver \href{https://www.ecmwf.int/en/forecasts/documentation-and-support/evolution-ifs/cycle-38r1-summary-changes}{ECMWF - OpenIFS version cy38r1nc4}, Community Emission Data System (CEDS) historical emission inventory \citep{GMD:Hoesly2018}

@@

The original dry deposition scheme of the OsloCTM2 and OsloCTM3 was not based on actual calculation of the deposition velocities from equations taking the state of the atmosphere into account, but rather paramerterized following the original work of \citet{AE:Wesely1989} with paramerter updates from \citet{JGR:Hough1991}. This means that seasonal day and night average deposition velocities for different land use types (water, forest, grass, tundra/desert, and ice and snow) had been used. Day was distiquished from night by solar zenith angles below 90\,\unit{^\circ}. Winter was defined by temperatures below 273.15\,\unit{K} for gridboxes containing land masses. Since winter and summer parameters for ocean are equal in this parameterization, no distinquct treatment was needed for ocean gridboxes. In addition, reduced uptake due to snow cover above 1\,\unit{m} and 10\,\unit{cm} for forest and grass/tundra, respectively, was taken into account. In the following, we will refer to this static parameterization as \emph{Wesley scheme}.\\

Getting away from the Wesely schme, we follow the Convention on Long-Range Transboundary Air Pollution (CLRTAP) implemented in the European Monitoring and Evaluation Programme (EMEP) MSC-W model \citep{Simpson2003,ACP:Simpson2012}. We will refere to this as \emph{EMEP scheme} thoughout the rest of the paper. The EMEP scheme is a more physical approach compared to the previous used Wesely scheme. It is used for the gaseous species \chem{O_3}, \chem{H_2O_2}, \chem{NO_2}, \chem{PAN}, \chem{SO_2}, \chem{NH_3}, \chem{HCHO}, and \chem{CH_3CHO}. Since \chem{CO} has a very small uptake and is not included in \citet{Simpson2003,ACP:Simpson2012}, the old OsloCTM2 \citep{OsloCTM2} parameterization is kept. In adition to the gaseous species, some of the aerosol deposition rates have also been modified, namely black carbon (BC) and organic carbon (OC), sulphuric (\chem{SO_4}, \chem{MSA}), and secondary organic aerosols (SOA). As displayed in Eq.~(\ref{eq:drydep_velo}), the dry deposition computation is typically splitted into three different contributions (aerodynamic $R_a$, quasi-laminar layer $R^i_b$, and canopy resistances $R^i_c$) which will be addressed seperately in the following.

\subsubsection{Aerodynamic resistance}
\label{subsubsec:Ra}
The aerodynamical resistance $R_a$ in \citet{Simpson2003,ACP:Simpson2012} is described as
\begin{equation}
  R_a = \frac{1}{\kappa u_*}\left[{\ln{\left(\frac{z-d}{z_0}\right)}-\Psi_m\left(\frac{z-d}{z_0}\right)-\Psi_m\left(\frac{z_0}{L}\right)}\right],
\end{equation}
with $\kappa$ the K\'{a}rm\'{a}n constant, $u_*$ the friction velocity, $\Psi_m$ the integrated stability equations for momentum, $d$ a constant (typically 0.7\,\unit{m}), and $L$ the Obukhov length. For certain values of $z$, $z_0$, and $L$, this may result in unphysical (negative) values for $R_a$. For this reason, we diverge from the EMEP scheme at this point and follow the method of \citet{Monteith1973}.
The sensible heat flux $\Phi_{Q_\text{sensible}}$ can be written as
\begin{equation}
\Phi_{Q_\text{sensible}} = \rho_\text{air} c_P \cdot \frac{\partial_z T}{\partial_z R_a},
\end{equation}
wherein $\rho_\text{air}$ is the air density, $c_P$ the specific heat at constant pressure $P$, $\partial_z T$ is the difference between the surface temperature $T_0$ and the temperature at reference height $T_z$, and $\partial_z R_a$ the diffusion resistance to sensible heat between surface ($z_0 = 0$) and the reference height $z$. From the eddy diffusion theorem, a similar formulation can be derived
\begin{equation}
  \Phi_{Q_\text{sensible}} = \rho_\text{air} c_P \cdot \frac{\partial_z T}{\partial_z u} \cdot u_*^2.
\end{equation}
From this and assuming $\partial_z R_a \rightarrow R_a $ and $\partial_z u \rightarrow u_z $ for a finite $z$, we find 
\begin{equation}
  R_a = \frac{u_z}{u_*^2}.
\end{equation}
As reference height $z$ we use the height at midlevel of the lowermost model level (roughly 8\,\unit{m}) and $u_z$ and $u_*$ are available from the meteorological input fields.

\subsubsection{Quasi-laminar layer resistance}
\label{subsubsec:Rb}
$R_b^i$ is species specific, as already mentioned, and its definition differs over land and ocean. In case a gridbox contains both land and ocean, a weighted mean is calculated using the land and ocean conductances, respectively.
\subsection*{Land}
Over land we use
\begin{equation}
  R_b^i = \frac{2}{\kappa u_*} \cdot \left(\frac{\text{Sc}_i}{\text{Pr}}\right)^{\frac{2}{3}},
\end{equation}
wherein $\text{Pr}$ is the Prandtl number (typically 0.72 for air and other gases) and $\text{Sc}_i$ is the Schmidt number for a gas $i$. From the definition of $\text{Sc}_i = \nu/D_i$, with the kinematic viscosity of air $\nu$, $\text{Sc}_i $ can be derived as water equivalent:
\begin{equation}
  \text{Sc}_i = \frac{\nu}{D_i} = \text{Sc}_\chem{H_2O}\cdot\frac{D_\chem{H_2O}}{D_i},
\end{equation}
with the molecular diffusivity for any gas $D_i$, the Schmidt number of water $\text{Sc}_\chem{H_2O}=0.6$ and a its molecular diffusivity $D_\chem{H_2O}=0.21\cdot10^{-4}\,\unit{m^2s^{-1}}$. The ratios $D_\chem{H_2O}/D_i$ are given in Table~\ref{tab:diffusivity}.
%
\begin{table}[t]
  \caption{Molecular diffusivity for a gas $i$ given as ration $D_\chem{H_2O}/D_i$.}
  \begin{tabular}{lc}
    \tophline
     & $D_\chem{H_2O}/D_i$\\
    \middlehline
    \chem{O_3} & 1.6\\
    \chem{SO_2} & 1.9\\
    \chem{NO_2} & 1.6\\
    \chem{H_2O_2} & 1.4\\
    \chem{HCHO} & 1.3\\
    \chem{NO} & 1.3\\
    \chem{CH_3CHO} & 1.6\\
    \bottomhline
  \end{tabular}
  %\belowtable{} % Table Footnotes
  \label{tab:diffusivity}
\end{table}
%
\subsection*{Ocean}
Over ocean we use
\begin{equation}
  R_b^i = \frac{1}{\kappa u_*}\cdot\ln\left({\frac{z_0}{D_i}\cdot \kappa u_*}\right)
\end{equation}
with a lower threashold of 10\,\unit{ms^{-1}} and an upper limit of 1000\,\unit{ms^{-1}}. The computation of roughness length $z_0$ over ocean is devided into a \emph{calm} and a \emph{rough} sea case, where windspeeds above 3\,\unit{ms^{-1}} are considered as boundary between both cases. Calm sea follows \citet{Hinze1975,Garrat1992} with a slightly increased coefficient:
\begin{equation}
  z_0^\text{calm} = \text{min}\left(2\cdot10^{-3}, 0.135 \frac{\nu}{u_*}\right).
\end{equation}
The kinematic viscosity of air $\nu = \mu/\rho$ can be computed from the absolute viscosity of air $\mu$, and substituting $\rho$ uising the gas equation by surface pressure $P_0$, 2\,\unit{m} temperature $T_\text{2M}$, and the universal gas constant for air $R_\text{air}$
\begin{equation}
  \nu = \frac{6.2\cdot 10^{-8}\,\unit{kg \cdot m^{-1}\cdot s^{-1} \cdot T^{-1}} \cdot T_\text{2M}}{\frac{P_0}{T_\text{2M}\cdot R_\text{air}}}.
\end{equation}
The rough sea case follows the method of \citet{Charnock1955,Wu1980}:
\begin{equation}
  z_0^\text{rough} = \text{min}\left(2\cdot10^{-3}, 0.018 \frac{u^2_*}{g}\right).
\end{equation}
The allowed maximum roughness length in both cases is set to 2\,\unit{mm}. Since the $z_0$ computed with this paramterization is very small, the resulting $R^i_b$ is often defined by its imposed lower limit of 10\,\unit{sm^{-1}}.

\subsubsection{Surface resistance}
\label{subsubsec:Rc}
%\begin{figure}[t]
%  \centering
%  \includegraphics[width=8.3cm]{pictures/pft_categories.pdf}\\
%  \includegraphics[width=8.3cm]{pictures/pft_categories_zonal.pdf}
%\caption{Plant functional types (PFT). (a) CML2 dynamic landuse 0.5x0.5\,\unit{^\circ}}
%\label{fig:pft_landuse}
%\end{figure}
Mapping of pft landuse types from either ISLSCP2 MODIS or CML2 categories to 9 classical biosphere types used in the EMEP scheme. EMEP categories: Forests, Mediterranean scrub, Crops, Moorland (savanna++), Grassland, Wetlands, Tundra, Desert, Water, Urban
Weighting of conductance with landuse fraction and sum. Inverse of this is gridbox average resistance.
-> \citep{ICP:MappingManual2017}

\subsection{Greening season and photosynthetic photon flux density}
\label{subsec:pre-pro}
      {\bf TODO: The neccessary pre-processing routines...}

@@

\subsection{Comparison with recent multi-model comparison}
\label{subsec:model}
Compare to \citet{ACP:Hardacre2015} as well as \citet{ACP:Luhar2017}.

\subsection{Comparison with observations}
\label{subsec:obs}
The sites used in \citet{ACP:Hardacre2015} and additional ozone zonal data from satellites and sondes (\url{http://bodekerscientific.com/}).

@@

%\appendix
%\section{}    %% Appendix A

%\subsection{}     %% Appendix A1, A2, etc.

>>>>

\begin{abstract}
 High concentrations of ozone in ambient air are hazardous not only to humans but to the ecosystem in general. The impact of ozone damage on vegetation and agricultural plants in combination with advancing climate change may affect food security in the future. While the future scenarios in themselves are uncertain, there are limiting factors constraining the accuracy of surface ozone modeling also at present: The distribution and amount of precursor/decomposing species, the stratosphere-troposphere exchange as well as scavenging processes. Removal of any gas through gravitational settling or by uptake by plants and soil is referred to as dry deposition. The process of dry deposition is important for predicting surface ozone concentrations and understanding the observed amount and increase of tropospheric background ozone. The conceptual dry deposition velocities are calculated following a resistance-analogous approach wherein aerodynamic, quasi laminar, and canopy resistances are key components, but these are hard to measure explicitly. In this paper, we present the update from an empirical dry deposition parameterization to a more process based implementation in the OsloCTM3. Our focus lies mainly on a stomatal conductance-based description of the canopy resistance. We evaluate the resulting modeled ozone dry deposition with respect to observations and multi-model studies. We will also estimate the impact of a set of parameters on the modeled ozone concentrations, both at the surface and in the free troposphere. We show that ozone dry deposition decreases with respect to the previous version, leading to an increase in surface ozone of up to ??\,\unit{\%}. {\bf TODO: Finish the results section as soon as there are results.}
\end{abstract}

@@

In the troposphere and in particular in ambient air, ozone is considered as a highly toxic pollutant. Since the industrial revolution, tropospheric background ozone concentrations have been increasing in the northern hemisphere. In recent years, the number of episodes of peak concentrations has been decreasing in North America and Europe due to the implementation of air quality regulations. At the same time, fast developing countries, like e.g., China or India, saw a significant increase in ozone related air pollution. Continuously high concentrations of ambient air ozone are hazardous to the whole ecosystem. It is estimated that ozone is cause to an increase in pre-mature deaths \citep{WHO2008}, an average global loss of yield in the four major crops (wheat, rice, maize, and soybean) of about 3--15\,\unit{\%} \citep{PJ:Ainsworth2017} as well as 7\,\unit{\%} loss in primary production in forestry \citep{GCB:Wittig2009,EP:Matyssek2012}. The impact of ozone damage on vegetation and agricultural plants may affect food security in the future especially in Asia \citep{GCB:Tang2013,NCC:Tai2014,AE:Chuwah2015} and might be an important additional feedback to climate change \citep{Nat:Sitch2007}.\\

@@

Removal of any substance through gravitational settling or by uptake by plants and soil is referred to as dry deposition. The process of dry deposition is important for predicting surface ozone concentrations and understanding the observed amount and increase of tropospheric background ozone. It is estimated that about $1000 \pm 200$\,\unit{Tg a^{-1}} of ozone are removed from the atmosphere by dry deposition processes \citep{ACP:Monks2015}. Conceptually, dry deposition is a product between surface ozone concentration \chem{[O_3](h_0)} and a dry deposition velocity $v^\chem{O_3}_d$. Dry deposition velocities $v^i_d$ (also referred to as conductance $G^i$) for any gaseous species $i$ are typically calculated following a resistance-analogous approach

@@

wherein aerodynamic $R_a$, quasi-laminar layer $R^i_b$, and canopy resistances $R^i_c$ are key components \citep{AE:Wesely1989}. For all gases $R_a$ is the same, while $R^i_b$ and $R^i_c$ vary from gas to gas and also depend on land surface types (e.g., ice/snow, water, urban, desert, agricultural land, deciduous forest, coniferous forest etc.). Originally, \citet{AE:Wesely1989} used fixed seasonal average dry deposition resistances for each land surface type. For all three types of resistances in this Wesely-type parameterization, more process-oriented formulations have been developed and validated over the years. \citet{ACP:Luhar2017} have validated ozone dry deposition to the ocean with respect to three different formulations of surface resistances. An update on the ozone surface resistance over snow and ice covered surfaces has been provided from combined model and observation studies \citep[e.g., $v^\chem{O_3}_\text{ice/snow} = 1/10000\,\unit{cm s^{-1}}$,][]{ACP:Helmig2007}. Canopy conductance is formulated at the single-leave level (stomatal conductance) for various plant function types (PFT) as well as single species based on empirical studies \citep{PTRS:Jarvis1976, BallBerry1987, ACP:Simpson2012, ICP:MappingManual2017}. But progress has also been made on process oriented modeling of stomatal conductance \citep{PP:Buckley2017}. The variety of differing formulations and choices of parameters leads to a wide spread in model intercomparisons \citep{ACP:Hardacre2015} and about 20\,\unit{\%} uncertainty on the resulting total dry deposition \citep{ACP:Monks2015}.\\

@@

\section{Model description}
\label{sec:model_des}
The OsloCTM3 is a three dimensional global chemistry transport model (CTM). The key components of the OsloCTM3 have been described and evaluated by \citet{GMD:Sovde2012}. A detailed account of the capabilities of the OsloCTM3 in simulating anthropogenic aerosol forcing in the past and recent past using the Community Emission Data System (CEDS) historical emission inventory \citep{GMD:Hoesly2018} is given by \citet{GMD:Lund2018}. \emph{The OsloCTM3 can also be coupled to MEGAN \citep{ACP:Guenther2006}. A publication focusing on this is planed.}
Technical: Partly parallelized (low number of effective cores 16, huge memory consumption if run on T159 resolution, code has been ported to fortran90 keeping the dynamical core UCI-CTM mainly intact.)
historical anthropogenic and biomass burning emissions \citep{ACP:Lamarque2010}, meteorological driver \href{https://www.ecmwf.int/en/forecasts/documentation-and-support/evolution-ifs/cycle-38r1-summary-changes}{ECMWF - OpenIFS version cy38r1nc4}, Community Emission Data System (CEDS) historical emission inventory \citep{GMD:Hoesly2018}

@@

In the original dry deposition scheme of the OsloCTM3, the state of the atmosphere was not taken into account. Dry deposition velocities were rather parameterized following the work of \citet{AE:Wesely1989} with parameter updates from \citet{JGR:Hough1991}. This means that seasonal day and night average deposition velocities for different land use types (water, forest, grass, tundra/desert, and ice and snow) were in use. Day was distinguished from night by solar zenith angles below 90\,\unit{^\circ}. Winter was defined by temperatures below 273.15\,\unit{K} for gridboxes containing land masses. Since winter and summer parameters for ocean are equal in this parameterization, no distinctive treatment was needed for ocean gridboxes. In addition, a reduced uptake due to snow cover above 1\,\unit{m} for forest and 10\,\unit{cm} for grass/tundra, respectively, was taken into account. In the following, we will refer to this static parameterization as \emph{Wesely scheme}.\\

For the new dry deposition scheme, we follow the Convention on Long-Range Transboundary Air Pollution (CLRTAP) implemented in the European Monitoring and Evaluation Programme (EMEP) MSC-W model \citep{WASP:Simpson2003,ACP:Simpson2012}. We will refer to this as \emph{EMEP scheme} throughout the rest of the paper. The EMEP scheme is a more physical approach compared to the previously used Wesely scheme, because it takes state (e.g., pressure, temperature) as well as dynamics (e.g., wind stress) of the atmosphere into account and also acknowledges to a certain degree the global variety of plants and the variability throughout the seasons. It is implemented for the gaseous species \chem{O_3}, \chem{H_2O_2}, \chem{NO_2}, \chem{PAN}, \chem{SO_2}, \chem{NH_3}, \chem{HCHO}, and \chem{CH_3CHO}. Since \chem{CO} has a very small uptake and is not included in \citet{WASP:Simpson2003,ACP:Simpson2012}, the Wesely parameterization is kept. In addition to the gaseous species, some of the aerosol deposition rates are also modified, namely black carbon (BC) and organic carbon (OC), sulfuric (\chem{SO_4}, \chem{MSA}), and secondary organic aerosols (SOA). As displayed in Eq.~(\ref{eq:drydep_velo}), the dry deposition computation is typically split into three different contributions (aerodynamic $R_a$, quasi-laminar layer $R^i_b$, and canopy resistances $R^i_c$) which will be addressed separately in the following.

\subsubsection{Aerodynamic resistance}
\label{subsubsec:Ra}
The aerodynamic resistance $R_a$ in \citet{WASP:Simpson2003,ACP:Simpson2012} is described as
\begin{equation}
  R_a = \frac{1}{\kappa u_*}\left[{\ln{\left(\frac{z-d}{z_0}\right)}-\Psi_m\left(\frac{z-d}{z_0}\right)-\Psi_m\left(\frac{z_0}{L}\right)}\right],
\end{equation}
with $\kappa$ the K\'{a}rm\'{a}n constant, $u_*$ the friction velocity, $\Psi_m$ the integrated stability equations for momentum, $d$ a constant (typically 0.7\,\unit{m}), and $L$ the Obukhov length. For certain values of $z$, $z_0$, and $L$, this may result in nonphysical (negative) values for $R_a$. For this reason, we diverge from the EMEP scheme at this point and follow the method of \citet{Monteith1973}.
The sensible heat flux $\Phi_{Q_\text{sensible}}$ can be written as
\begin{equation}
  \Phi_{Q_\text{sensible}} = \rho_\text{air} c_P \cdot \frac{\partial_z T}{\partial_z R_a},
  \label{eq:sens_heat}
\end{equation}
wherein $\rho_\text{air}$ is the air density, $c_P$ the specific heat at constant pressure $P$, $\partial_z T$ is the difference between the surface temperature $T_0$ and the temperature at reference height $T_z$, and $\partial_z R_a$ the diffusion resistance to sensible heat between surface ($z_0 = 0$) and the reference height $z$. From the eddy diffusion theorem, a similar formulation can be derived
\begin{equation}
  \Phi_{Q_\text{sensible}} = \rho_\text{air} c_P \cdot \frac{\partial_z T}{\partial_z u} \cdot u_*^2.
  \label{eq:eddy_theo}
\end{equation}
Comparing Eq.~(\ref{eq:sens_heat}) and Eq.~(\ref{eq:eddy_theo}) and assuming $\partial_z R_a \rightarrow R_a $ and $\partial_z u \rightarrow u_z $ for a finite $z$, we find 
\begin{equation}
  R_a = \frac{u_z}{u_*^2}.
\end{equation}
As reference height $z$, we use the height at mid-level of the lowermost model level (roughly 8\,\unit{m}), while $u_z$ and $u_*$ are available from the meteorological input fields.

\subsubsection{Quasi-laminar layer resistance}
\label{subsubsec:Rb}

$R_b^i$ is species specific, as already mentioned, and its definition differs over land and ocean. In case a gridbox contains both land and ocean, a weighted mean is calculated using the land and ocean conductances, respectively.
\subsection*{Land}
Over land we use the definition \citep[Eq.~(53),][]{ACP:Simpson2012}
\begin{equation}
  R_b^i = \frac{2}{\kappa u_*} \cdot \left(\frac{\text{Sc}_i}{\text{Pr}}\right)^{\frac{2}{3}},
\end{equation}
wherein $\text{Pr}$ is the Prandtl number (typically 0.72 for air and other gases) and $\text{Sc}_i$ is the Schmidt number for a gas $i$. From the definition of $\text{Sc}_i = \nu/D_i$, with the kinematic viscosity of air $\nu$, $\text{Sc}_i $ can be derived as water equivalent:
\begin{equation}
  \text{Sc}_i = \frac{\nu}{D_i} = \text{Sc}_\chem{H_2O}\cdot\frac{D_\chem{H_2O}}{D_i},
\end{equation}
with the molecular diffusivity for any gas $D_i$, the Schmidt number of water $\text{Sc}_\chem{H_2O}=0.6$ and a its molecular diffusivity $D_\chem{H_2O}=0.21\cdot10^{-4}\,\unit{m^2s^{-1}}$. The ratios $D_\chem{H_2O}/D_i$ are given in Table~\ref{tab:diffusivity}.
%
\begin{table}[t]
  \caption{Molecular diffusivity for a gas $i$ given as ration $D_\chem{H_2O}/D_i$, reactivity $f^i_0$, and effective Henry's Law constant $H^i_*$.}
  \begin{tabular}{lcrr}
    \tophline
    & $D_\chem{H_2O}/D_i$ & $f^i_0$ & $H^i_*$\\
    & & & (\unit{M\cdot atm^{-1}})\\
    \middlehline
    \chem{O_3}     & 1.6 & 1.0 & $10^{-2}$\\
    \chem{SO_2}    & 1.9 & 0.0 & $10^5$\\
    \chem{NO_2}    & 1.6 & 0.1 & $10^{-2}$\\
    \chem{H_2O_2}  & 1.4 & 1.0 & $10^5$\\
    \chem{HCHO}    & 1.3 & 0.0 & $6 \cdot 10^3$\\
    \chem{NO}      & 1.3 & 0.0 & $2 \cdot 10^{-3}$\\
    \chem{CH_3CHO} & 1.6 & 15.0 & $0.0$\\
    \bottomhline
  \end{tabular}
  %\belowtable{} % Table Footnotes
  \label{tab:diffusivity}
\end{table}
%
\subsection*{Ocean}
Over ocean we use the definition \citep[Eq.~(54),][]{ACP:Simpson2012}
\begin{equation}
  R_b^i = \frac{1}{\kappa u_*}\cdot\ln\left({\frac{z_0}{D_i}\cdot \kappa u_*}\right)
\end{equation}
with an imposed a lower threshold of 10\,\unit{sm^{-1}} and an upper limit of 1000\,\unit{sm^{-1}}. The computation of roughness length $z_0$ over ocean is divided into a \emph{calm} and a \emph{rough} sea case, where wind speeds above 3\,\unit{ms^{-1}} are considered as boundary between both cases. Calm sea follows \citet{Hinze1975,Garratt1992} with a slightly increased coefficient:
\begin{equation}
  z_0^\text{calm} = \text{min}\left(2\cdot10^{-3}, 0.135 \cdot \frac{\nu}{u_*}\right).
\end{equation}
The kinematic viscosity of air $\nu = \mu/\rho$ can be computed from the absolute viscosity of air $\mu$, and substituting $\rho$ with surface pressure $P_0$ using the ideal gas law, 2\,\unit{m} temperature $T_\text{2M}$, and the universal gas constant for air $R_\text{air}$
\begin{equation}
  \nu = \frac{T_\text{2M}}{\frac{P_0}{T_\text{2M}\cdot R_\text{air}}} \cdot 6.2\cdot 10^{-8}\,\unit{kg m^{-1} s^{-1} K^{-1}}.
\end{equation}
The rough sea case follows the method of \citet{QJRMS:Charnock1955,JPO:Wu1980}:
\begin{equation}
  z_0^\text{rough} = \text{min}\left(2\cdot10^{-3}, 0.018 \cdot \frac{u^2_*}{g}\right).
\end{equation}
The allowed maximum roughness length in both cases is set to 2\,\unit{mm}. Since the $z_0$ computed with this parameterization is very small, the resulting $R^i_b$ is often defined by its imposed lower limit of 10\,\unit{sm^{-1}}.

\subsubsection{Surface resistance}
\label{subsubsec:Rc}

The surface resistance $R_c^i$ consists of both, stomatal and non-stomatal resistances. When computing gridbox averages, we use the sum of conductances (deposition velocities)
\begin{equation}
  G_c^i = \text{LAI} \cdot G_\text{sto}^i + G_\text{ns}^i, 
\end{equation}
wherein $\text{LAI}$ is the leaf area index (zero for non-vegetated surfaces), $G_\text{sto}^i$ the stomatal conductance, and $G_\text{ns}^i$ the non-stomatal conductance.

{\bf TODO: Mapping of pft land surface types from either ISLSCP2 MODIS or CML2 categories to 9 classical biosphere types used in the EMEP scheme. EMEP categories: Forests, Mediterranean scrub, Crops, Moorland (savanna++), Grassland, Wetlands, Tundra, Desert, Water, Urban. Weighting of conductance with land surface fraction and sum. Inverse of this is gridbox average resistance.}

\subsubsection*{Stomatal conductance}
-> \citep{ICP:MappingManual2017}
{\bf TODO: Write about latitude dependent vegetation height and EMEP algorithm in general.}

\subsubsection*{Non-stomatal conductance}
In the EMEP scheme, non-stomatal conductances are explicitly calculated for \chem{O_3}, \chem{SO_2}, \chem{HNO_3}, and \chem{NH_3}. For all other species, an interpolation between \chem{O3} and \chem{SO_2} values is carried out.

The non-stomatal conductance for \chem{O_3} consists of two terms, one depending on vegetation type and one depending in the soil/surface. For a land surface type $N$, we can write
\begin{equation}
  G_\text{ns}^\chem{O_3}(N) = \frac{\text{SAI}(N)}{r_\text{ext}} + \frac{1}{R_\text{inc}(N) + R_\text{gs}^\chem{O_3}(N)}.
\end{equation}
$\text{SAI}(N)$ is the surface area index for vegetation type $N$, which is $\text{LAI}$ plus some value representing cuticles and others surfaces. The external leaf resistance $r_\text{ext} = 2000\,\unit{sm^{-1}}\cdot F_T$. Below $-1\,\unit{^\circ C}$,s a temperature correction factor $F_T$ is used
\begin{equation}
  F_T = \exp{(-2\cdot(1+T_0))}.
\end{equation}
It ranges between 1--2.
\subsection{Pre-processing}
\label{subsec:pre-pro}
      {\bf TODO: The necessary pre-processing routines...}

      \subsubsection{Greening season}

      \subsubsection{Photosynthetic photon flux density (PPFD)}

@@

\subsection{Comparison with recent multi-model comparison}
\label{subsec:model}
Compare to \citet{ACP:Hardacre2015} as well as \citet{ACP:Luhar2017}. \citep{AE:Derwent2018} Uncertainties in surface ozone sources and sinks larger than in formulations? -> Need to read this one...

\subsection{Comparison with observations}
\label{subsec:obs}
The sites used in \citet{ACP:Hardacre2015} and additional ozone zonal data from satellites and sondes (\url{http://bodekerscientific.com/}). \citep{AE:Hu2017} -> need to look into this one...

@@

\appendix
\section{Land use and plant function types (PFT)}    %% Appendix A
\appendixfigures
\begin{figure}[t]
  \centering
  \includegraphics[width=8.3cm]{pictures/pft_categories.pdf}\\
  \includegraphics[width=8.3cm]{pictures/pft_categories_zonal.pdf}
  \caption{Plant functional types (PFT). (a) CML2 dynamic land surface 0.5x0.5\,\unit{^\circ}}
  \label{fig:pft_landsurface}
\end{figure}
\subsection{}     %% Appendix A1, A2, etc.
